A new simplified manual tour, with examples in mathematica

### 2022-04-05 paper/paper.Rmd
done with writing out the algorithms

---
title: |
  A new simplified manual tour, with examples in mathematica
type: Short Technical Note
author:
  - name: Alex Aumann
    affil: a
    email: aaum0002@student.monash.edu
  - name: German Valencia
    affil: a
    email: german.valencia@monash.edu
  - name: Ursula Laa
    affil: b
    email: ursula.laa@boku.ac.at
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Something here
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
output: rticles::tf_article
keep_tex: true
---

@@

@@

Describe the manip space, out of manual tour

## New definition

1. Provide a k-D projection
2. Select variable to control
3. Detect or provide direction of change
4. Update projection, and orthonormalise

Need to think about how the values for selected variable are made exact, that is constrained orthonormalisation.

Need to think about checks, and error catching, maybe in implementation section. If selected variable has 0 coefficient, will that generate an orthonormalisation error?

```{r manualsequence, outwidth="80%", width=12, height=8, fig.cap="Sequence of projections where V3 contribution is changed."}
source("../code/manual.R")

# Base plot set up
plot.new()
par(pty="s", xaxt="n", yaxt="n", bty="n",
      omi=c(0,0,0,0), mar=c(0,0,0,0),
    mfrow=c(3,4))

@@

# Now set a new position
vchange <- 3
eps <- 0.02
plot_basis(A, vchange)

# Iterate and save images
for (i in 1:11) {
  flname <- paste0("mtour_plots/mtour", i, ".png")
  A[vchange,] <- A[vchange,]*(1-i*eps)
  A <- tourr::orthonormalise(A)
  #png(flname, width=250, height=250)
  plot_basis(A, vchange)
  #dev.off()
}
```

>>>>

---
title: |
  A new simplified manual tour, with examples in mathematica
type: Short Technical Note
author:
  - name: Alex Aumann
    affil: a
    email: aaum0002@student.monash.edu
  - name: German Valencia
    affil: a
    email: german.valencia@monash.edu
  - name: Ursula Laa
    affil: b
    email: ursula.laa@boku.ac.at
  - name: Dianne Cook
    affil: c
    email: dicook@monash.edu
affiliation:
  - num: a
    address: |
      School of Physics and Astronomy, Monash University
  - num: b
    address: |
      Institute of Statistics, University of Natural Resources and Life Sciences, Vienna
  - num: b
    address: |
      Department of Econometrics and Business Statistics, Monash University
bibliography: biblio.bib
geometry: margin=2.5cm
abstract: |
  Something here
keywords: |
  data visualisation; grand tour; statistical computing; statistical graphics; multivariate data; dynamic graphics
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \usepackage{setspace}
  \usepackage{graphicx}
output: rticles::tf_article
keep_tex: true
---

@@

An orthonormal basis ($A_{p\times d}$) and a variable id ($m \in \{1, ..., p\}$) to control are provided to initialise a manual tour. A method to update the values of the component of the controlled variable $V_m$ is needed.

@@

In the original work, the method for updating component values, for a 2D projection, was built trackball controls in 3D. A 3D manipulation space is created, as illustrated in Figure \ref{manipspace}, where the controlled variable has full range of motion from -1 to 1. Movements of a cursor are recorded and converted into changes in the values of $V_m$ to change it's values in the displayed 2D projection. Movement could also be constrained to be only in horizontal, vertical, radial or angular motions.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.8\textwidth]{figures/manip_space.pdf}}
\caption{Original contruction of the manual tour designed for 2D projections and created a 3D space from which to utilise track ball controls to change it's contribution. (Figure 3 from Cook and Buja (1997).)}
\label{manipspace}
\end{figure*}

## New simpler definition

The new approach emerged from experiments in mathematica. The components corresponding to $V_m$ are directly controlled by cursor movement, which updates row $m$ of $A$. The updated matrix is then orthonormalised.

### Algorithm 1

1. Provide $A$, and $m$. (Note that $m$ could also be automatically chosen as the component that is closest to the cursor position.)
2. Change values in row $m$, giving $A^*$. A large change in these values would correspond to making a large jump from the current projection. Small changes would correspond to tracking a cursor, making small jumps from the current projection.
3. Orthonormalise $A^*$, using Gram-Schmidt. For $d=2$, and $A^* = \left[ {\boldmath a}_{.1}~{\boldmath a}_{.2}\right]$, the steps are:

    i. Normalise ${\boldmath a}_{.1}$, and ${\boldmath a}_{.2}$.
    ii. ${\boldmath a}^*_{.2} = {\boldmath a}_{.2} - {\boldmath a}_{.1}^T{\boldmath a}_{.2}{\boldmath a}_{.1}$.
    iii. Normalise ${\boldmath a}^*_{.2}$.

This algorithm will produce the changes to a projection as illustrated in Figure \ref{fig:manualsequence} (top row). The controlled variable, $V_m$, corresponds to the black line, and sequential changes to row $m$ of $A$ can be seen to roughly follow a specified position (orange dot). Changes in the other components happen as a result of the orthonormalisation, but are uncontrolled.

### Algorithm 2

The problem with Algorithm 1 is that the precise values for $V_m$ cannot be specified because the orthonormalisation wil change them. This modification will maintain the components of $V_m$ precisely (Figure \ref{fig:manualsequence} (bottom row)). The algorithm is as follows:

1. Provide $A$, and $m$.
2. Change values in row $m$, giving $A^*$.
3. Store row $m$ separately, and zero the values of row $m$ in $A^*$, giving $A^{*0}$. 
4. Orthonormalise $A^{*0}$, using Gram-Schmidt.
5. Replace row $m$ with the original values, giving $A^{**}$. 
6. For $d=2$, adjust the values of ${\boldmath a}^{**}_{.2}$ using

$$a^{**}_{j2}+\frac{a_{m1}a_{m2}}{p-1}, j=1, ..., p, j\neq m$$.

which ensures that

$$\sum_{j=1, j\neq m}^p a^{**}_{j1}a^{**}_{j2} + a_{m1}a_{m2} = 0$$.

If $d>2$ the process would be sequentially repeated in the same manner that Gram-Schmidt is applied sequentially to orthormalise the columns of a matrix. If $d=1$ no orthonormalisation is needed, and the projection vector would simply need to be normalised after each adjustment.

<!-- Need to think about how the values for selected variable are made exact, that is constrained orthonormalisation.

Need to think about checks, and error catching, maybe in implementation section. If selected variable has 0 coefficient, will that generate an orthonormalisation error?-->

```{r manualsequence, outwidth="80%", width=12, height=5, fig.cap="Sequence of projections where contribution of one variable is controlled (black) is changed: (top) unconstrained orthonormalisation, (bottom) constrained as specified. The dot (orange) indicates the chosen values for the controlled variable. For the constrained orthonormalisation it can be seen to precisely match the axis, but not so for the unconstrained orthonormalisation."}
source("../code/plot_basis.R")
source("../code/linear_alg.R")

# Base plot set up
plot.new()
par(pty="s", xaxt="n", yaxt="n", bty="n",
      omi=c(0,0,0,0), mar=c(0,0,0,0),
    mfrow=c(2,4))

@@

# Now set a new position
vchange <- 3
eps <- 0.1
plot_basis(A, vchange, vcol="black")

# Iterate and save images
# Unconstrained orthonormalisation
Anew <- A
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- tourr::orthonormalise(Anew)
  plot_basis(Anew, vchange, vcol="black")
  points(exact[1], exact[2], pch=16, col="orange")
}

# Iterate and save images
# Constrained orthonormalisation
Anew <- A
plot_basis(A, vchange, vcol="black")
for (i in 1:3) {
  Anew[vchange,] <- A[vchange,]*(1-i*eps)
  exact <- Anew[vchange,]
  Anew <- orthonormalise_frozen(Anew, vchange)
  plot_basis(Anew, vchange, vcol="black")
  points(exact[1], exact[2], pch=16, col="orange")
}
```

### 2022-04-05 paper/paper.tex
done with writing out the algorithms

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}

@@

@@

Describe the manip space, out of manual tour

\hypertarget{new-definition}{%
\subsection{New definition}\label{new-definition}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Provide a k-D projection
\item
  Select variable to control
\item
  Detect or provide direction of change
\item
  Update projection, and orthonormalise
\end{enumerate}

Need to think about how the values for selected variable are made exact,
that is constrained orthonormalisation.

Need to think about checks, and error catching, maybe in implementation
section. If selected variable has 0 coefficient, will that generate an
orthonormalisation error?

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/manualsequence-1.pdf}
\caption{Sequence of projections where V3 contribution is changed.}
\end{figure}

>>>>

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}
\usepackage{setspace}
\usepackage{graphicx}

@@

An orthonormal basis (\(A_{p\times d}\)) and a variable id
(\(m \in \{1, ..., p\}\)) to control are provided to initialise a manual
tour. A method to update the values of the component of the controlled
variable \(V_m\) is needed.

@@

In the original work, the method for updating component values, for a 2D
projection, was built trackball controls in 3D. A 3D manipulation space
is created, as illustrated in Figure \ref{manipspace}, where the
controlled variable has full range of motion from -1 to 1. Movements of
a cursor are recorded and converted into changes in the values of
\(V_m\) to change it's values in the displayed 2D projection. Movement
could also be constrained to be only in horizontal, vertical, radial or
angular motions.

\begin{figure*}[ht]
\centerline{\includegraphics[width=0.8\textwidth]{figures/manip_space.pdf}}
\caption{Original contruction of the manual tour designed for 2D projections and created a 3D space from which to utilise track ball controls to change it's contribution. (Figure 3 from Cook and Buja (1997).)}
\label{manipspace}
\end{figure*}

\hypertarget{new-simpler-definition}{%
\subsection{New simpler definition}\label{new-simpler-definition}}

The new approach emerged from experiments in mathematica. The components
corresponding to \(V_m\) are directly controlled by cursor movement,
which updates row \(m\) of \(A\). The updated matrix is then
orthonormalised.

\hypertarget{algorithm-1}{%
\subsubsection{Algorithm 1}\label{algorithm-1}}

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Provide \(A\), and \(m\). (Note that \(m\) could also be automatically
  chosen as the component that is closest to the cursor position.)
\item
  Change values in row \(m\), giving \(A^*\). A large change in these
  values would correspond to making a large jump from the current
  projection. Small changes would correspond to tracking a cursor,
  making small jumps from the current projection.
\item
  Orthonormalise \(A^*\), using Gram-Schmidt. For \(d=2\), and
  \(A^* = \left[ {\boldmath a}_{.1}~{\boldmath a}_{.2}\right]\), the
  steps are:

  \begin{enumerate}
  \def\labelenumii{\roman{enumii}.}
  \tightlist
  \item
    Normalise \({\boldmath a}_{.1}\), and \({\boldmath a}_{.2}\).
  \item
    \({\boldmath a}^*_{.2} = {\boldmath a}_{.2} - {\boldmath a}_{.1}^T{\boldmath a}_{.2}{\boldmath a}_{.1}\).
  \item
    Normalise \({\boldmath a}^*_{.2}\).
  \end{enumerate}
\end{enumerate}

This algorithm will produce the changes to a projection as illustrated
in Figure \ref{fig:manualsequence} (top row). The controlled variable,
\(V_m\), corresponds to the black line, and sequential changes to row
\(m\) of \(A\) can be seen to roughly follow a specified position
(orange dot). Changes in the other components happen as a result of the
orthonormalisation, but are uncontrolled.

\hypertarget{algorithm-2}{%
\subsubsection{Algorithm 2}\label{algorithm-2}}

The problem with Algorithm 1 is that the precise values for \(V_m\)
cannot be specified because the orthonormalisation wil change them. This
modification will maintain the components of \(V_m\) precisely (Figure
\ref{fig:manualsequence} (bottom row)). The algorithm is as follows:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  Provide \(A\), and \(m\).
\item
  Change values in row \(m\), giving \(A^*\).
\item
  Store row \(m\) separately, and zero the values of row \(m\) in
  \(A^*\), giving \(A^{*0}\).
\item
  Orthonormalise \(A^{*0}\), using Gram-Schmidt.
\item
  Replace row \(m\) with the original values, giving \(A^{**}\).
\item
  For \(d=2\), adjust the values of \({\boldmath a}^{**}_{.2}\) using
\end{enumerate}

\[a^{**}_{j2}+\frac{a_{m1}a_{m2}}{p-1}, j=1, ..., p, j\neq m\].

which ensures that

\[\sum_{j=1, j\neq m}^p a^{**}_{j1}a^{**}_{j2} + a_{m1}a_{m2} = 0\].

If \(d>2\) the process would be sequentially repeated in the same manner
that Gram-Schmidt is applied sequentially to orthormalise the columns of
a matrix. If \(d=1\) no orthonormalisation is needed, and the projection
vector would simply need to be normalised after each adjustment.

\begin{figure}
\centering
\includegraphics{paper_files/figure-latex/manualsequence-1.pdf}
\caption{Sequence of projections where contribution of one variable is
controlled (black) is changed: (top) unconstrained orthonormalisation,
(bottom) constrained as specified. The dot (orange) indicates the chosen
values for the controlled variable. For the constrained
orthonormalisation it can be seen to precisely match the axis, but not
so for the unconstrained orthonormalisation.}
\end{figure}
